---
title: "Getting Started with dcmtobids"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dcmtobids}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r libs, include = FALSE}
library(dcmtobids)
library(jsonlite)
```

## The problem

Your scanner hands you a pile of DICOMs with filenames like
`001_MR_ABCD_T1_MPRAGE_2024-01-15.dcm`. BIDS wants
`sub-01/anat/sub-01_T1w.nii.gz`. Doing that by hand is miserable and
error-prone; scripting it one participant at a time buries the rules in
ad-hoc code that rots between grants.

`dcmtobids` lets you describe the mapping *once* as a JSON config, then
plans and executes the conversion — with a dry run you can inspect, a
structured QA report you can gate CI on, and a post-hoc "rewire
IntendedFor" command for when your config changes after conversion has
already happened.

## A first conversion

We'll skip `dcm2niix` for this tour and work directly from the
NIfTI+JSON sidecars it would produce ("helper-style" inputs).
Everything below runs without real DICOM data installed.

```{r paths, include = FALSE}
workspace   <- file.path(tempdir(), paste0("dcmtobids-vig-", sample.int(1e6, 1)))
input_dir   <- file.path(workspace, "helper")
bids_dir    <- file.path(workspace, "bids")
config_path <- file.path(workspace, "config.json")
dir.create(input_dir, recursive = TRUE)
```

First, two helper sidecars — a T1 and a resting-state run — as
`dcm2niix` would have written them:

```{r make-helper}
writeLines(
  '{"SeriesDescription": "T1_MPRAGE"}',
  file.path(input_dir, "001_T1.json")
)
writeLines("placeholder", file.path(input_dir, "001_T1.nii.gz"))

writeLines(
  '{"SeriesDescription": "task-rest_bold", "RepetitionTime": 2.0}',
  file.path(input_dir, "002_rest.json")
)
writeLines("placeholder", file.path(input_dir, "002_rest.nii.gz"))
```

Then a minimal config with two rules — one per series we expect to see:

```{r write-config}
cfg <- list(
  descriptions = list(
    list(
      datatype = "anat",
      suffix   = "T1w",
      criteria = list(SeriesDescription = "*T1*")
    ),
    list(
      datatype = "func",
      suffix   = "bold",
      custom_entities = list("task-rest"),
      criteria = list(SeriesDescription = "*rest*")
    )
  )
)
writeLines(toJSON(cfg, auto_unbox = TRUE, pretty = TRUE), config_path)
```

And one call that plans + converts:

```{r first-conversion}
result <- run_conversion(
  config_path       = config_path,
  input_dir         = input_dir,
  bids_dir          = bids_dir,
  participant_label = "01"
)
result[c("converted", "skipped", "errors")]
```

Here's the BIDS tree that appeared:

```{r show-tree, echo = FALSE}
show_tree <- function(root, prefix = "") {
  entries <- sort(list.files(root, full.names = TRUE, include.dirs = TRUE,
                             all.files = FALSE, no.. = TRUE))
  entries <- entries[!grepl("/\\.dcmtobids$", entries)]
  for (i in seq_along(entries)) {
    is_last <- i == length(entries)
    cat(prefix, if (is_last) "`-- " else "|-- ", basename(entries[[i]]),
        "\n", sep = "")
    if (dir.exists(entries[[i]])) {
      show_tree(entries[[i]],
                prefix = paste0(prefix, if (is_last) "    " else "|   "))
    }
  }
}
cat(basename(bids_dir), "\n", sep = "")
show_tree(bids_dir)
```

Four files written, BIDS layout honored. The rest of this vignette
unpacks *how* that happened and what else you can do at each step.

## The mental model

A dcmtobids config is a **list of `descriptions`**. Each description is
a rule: *if a sidecar matches these `criteria`, write it to
`<datatype>/<suffix>` with these optional entities and sidecar edits.*

| Field              | Meaning                                                |
|:-------------------|:-------------------------------------------------------|
| `datatype`         | BIDS datatype (`anat`, `func`, `fmap`, `dwi`, …)       |
| `suffix`           | BIDS suffix (`T1w`, `bold`, `epi`, …)                  |
| `criteria`         | Required sidecar JSON fields (glob, regex, or numeric) |
| `custom_entities`  | BIDS key-value tokens (`task-rest`, `dir-AP`, …)       |
| `id`               | Short tag used by `IntendedFor` / `Sources`            |
| `sidecar_changes`  | Fields to add or overwrite in the output JSON          |

The pipeline itself has four verbs:

- `plan_conversion()` — match sidecars, return a data-frame plan
- `print_dry_run_plan()` — preview source → destination mappings
- `convert_plan()` — execute the plan (copy + rewrite sidecars)
- `summarize_plan()` — structured QA report with a gate (`pass`/`warn`/`fail`)

`run_conversion()` bundles the first three; `run_dicom_conversion()`
prepends a `dcm2niix` pass for you.

## Look before you leap: dry-run

Build a plan without touching the output directory:

```{r dry-run}
config <- read_config(config_path)
plan   <- plan_conversion(
  config            = config,
  input_dir         = input_dir,
  participant_label = "01"
)
plan[, c("source_json", "status", "destination_dir", "destination_stem")]
```

Each row is one helper sidecar, tagged `MATCHED`, `SKIPPED` (no rule
matched), or `AMBIGUOUS` (multiple rules matched). Only `MATCHED` rows
produce files when you later call `convert_plan()`; ambiguity and
missed matches are *surfaced*, never executed silently.

For a one-line-per-row view you'd see in the CLI:

```{r print-plan}
print_dry_run_plan(plan, bids_dir = bids_dir)
```

## QA report and CI gating

`summarize_plan()` produces a stable, machine-readable report:

```{r qa-report}
report <- summarize_plan(plan)
report$qa
```

`gate` summarizes severity:

- `pass` — no warning or error events
- `warn` — at least one warning (no-match, ambiguous, missing ID, …)
- `fail` — at least one error event

`codes` aggregates stable codes (`W_NO_MATCH`, `W_AMBIGUOUS_MATCH`,
`W_MISSING_ID_REFERENCE`, `E_CONVERSION_ERRORS`, …) so you can pattern-
match in automation. The CLI flag `--fail-on-warning` translates any
non-`pass` gate into a non-zero exit status, which is the standard way
to gate CI pipelines.

## Richer criteria

Beyond glob patterns, `criteria` supports list equality and numeric
ranges. Add a diffusion helper to see both:

```{r extra-helper}
writeLines(
  toJSON(list(
    SeriesDescription = "ep2d_diff",
    EchoTime          = 0.0738,
    ImageType         = list("ORIGINAL", "PRIMARY", "DIFFUSION", "ND")
  ), auto_unbox = TRUE),
  file.path(input_dir, "003_dwi.json")
)
writeLines("placeholder", file.path(input_dir, "003_dwi.nii.gz"))
```

A rule that requires both the exact `ImageType` list *and* an echo time
within a window:

```{r float-criteria}
dwi_rule <- list(
  datatype = "dwi",
  suffix   = "dwi",
  criteria = list(
    ImageType = list("ORIGINAL", "PRIMARY", "DIFFUSION", "ND"),
    EchoTime  = list(btwe = list(0.07, 0.08))
  )
)
```

Supported numeric operators are `gt`, `lt`, `ge`, `le`, `btw`
(exclusive) and `btwe` (inclusive). Use the `any` key to alternate
patterns against a single field:

```{r any-criteria, eval = FALSE}
criteria <- list(
  SeriesDescription = list(any = list("*rest*", "*resting*"))
)
```

Switch the whole config to regex rather than glob by setting
`search_method = "re"` at the config top level.

## Linking fieldmaps to functionals

BIDS's `IntendedFor` points a fieldmap at the functional images it
calibrates. In dcmtobids, you tag the *target* with an `id` and
reference it from the *source*'s `sidecar_changes`:

```{r fmap-config}
cfg_fmap <- list(
  descriptions = list(
    list(
      id              = "rest",
      datatype        = "func",
      suffix          = "bold",
      custom_entities = list("task-rest"),
      criteria        = list(SeriesDescription = "*rest*")
    ),
    list(
      datatype        = "fmap",
      suffix          = "epi",
      custom_entities = list("dir-AP"),
      criteria        = list(SeriesDescription = "*fmap*"),
      sidecar_changes = list(IntendedFor = list("rest"))
    )
  )
)
```

At conversion, dcmtobids resolves `"rest"` to the BIDS path of the
matching functional (`bids::sub-01/func/sub-01_task-rest_bold.nii.gz`
by default, or a participant-relative path if the config sets
`bids_uri = "relative"`).

## Repairing IntendedFor after the fact

Already ran the conversion and *then* noticed the fmap → bold link is
wrong? `populate_intended_for_plan()` rewrites only the `IntendedFor`
and `Sources` keys of existing BIDS sidecars — no file moves, no
re-copying NIfTIs, so it's safe to run repeatedly after editing the
config.

```{r populate, eval = FALSE}
config2 <- read_config(updated_config_path)
plan2   <- plan_conversion(config2, input_dir, participant_label = "01")
populate_intended_for_plan(plan2, bids_dir = bids_dir)
```

The matching CLI verb is `dcmtobids populate-intended-for`; it also
writes a run manifest (see below) so you have an audit trail of the
repair.

## CLI equivalents

Every R call has a CLI verb. `dcmtobids::install_cli()` drops a
launcher into `~/.local/bin`.

| R function                                   | CLI command                           |
|:---------------------------------------------|:--------------------------------------|
| `read_config()` / `validate_config()`        | `dcmtobids validate-config`           |
| `plan_conversion()` + `print_dry_run_plan()` | `dcmtobids dry-run`                   |
| `convert_plan()`                             | `dcmtobids convert`                   |
| `run_dicom_conversion()`                     | `dcmtobids run`                       |
| `populate_intended_for_plan()`               | `dcmtobids populate-intended-for`     |
| `inspect_sidecars()`                         | `dcmtobids inspect`                   |
| —                                            | `dcmtobids doctor` (env readiness)    |
| —                                            | `dcmtobids init-config` (starter)     |

Every CLI command that writes output also drops a **run manifest** at
`<bids_dir>/.dcmtobids/runs/<timestamp>-<cmd>-<nonce>/` containing the
plan, the config in use, the QA report, and the exact argv — an
auditable receipt of each conversion.

## Where to go next

- `?plan_conversion` — all planning options, including
  `auto_extract_entities` (task/dir/echo extraction from sidecar fields)
  and `do_not_reorder_entities` (keep custom entity order verbatim)
- `?convert_plan` — clobber semantics, post-op hooks,
  `bids_uri = "URI"` vs `"relative"`
- `?summarize_plan` — full QA report structure and warning-code list
- `inst/schema/config-schema.json` — the config JSON schema
- `dcmtobids help` at the shell for the full CLI surface

```{r cleanup, include = FALSE}
unlink(workspace, recursive = TRUE, force = TRUE)
```
