---
title: "Introduction to dcmtobids"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dcmtobids}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`dcmtobids` provides an R implementation of key DICOM-to-BIDS planning and
conversion steps:

- validate a config
- build a conversion plan (dry-run)
- execute conversion
- produce a structured QA report

This vignette uses precomputed helper-like files (`.json`, `.nii.gz`) so it can
run without real DICOM data or `dcm2niix`.

## Setup

```{r}
library(dcmtobids)
library(jsonlite)

td <- tempfile("dcmtobids-vig-")
dir.create(td)

input_dir <- file.path(td, "input")
bids_dir <- file.path(td, "bids")
config_path <- file.path(td, "config.json")

dir.create(input_dir)
dir.create(bids_dir)
```

## Create Minimal Example Inputs

```{r}
writeLines('{"SeriesDescription":"T1_MPRAGE"}', file.path(input_dir, "001.json"))
writeLines("dummy", file.path(input_dir, "001.nii.gz"))

writeLines('{"SeriesDescription":"task-rest_bold"}', file.path(input_dir, "002.json"))
writeLines("dummy", file.path(input_dir, "002.nii.gz"))
```

## Write and Validate Config

```{r}
cfg <- list(
  descriptions = list(
    list(
      datatype = "anat",
      suffix = "T1w",
      criteria = list(SeriesDescription = "*T1*")
    ),
    list(
      datatype = "func",
      suffix = "bold",
      custom_entities = list("task-rest"),
      criteria = list(SeriesDescription = "*rest*")
    )
  )
)

writeLines(toJSON(cfg, auto_unbox = TRUE, pretty = TRUE), config_path)

config <- read_config(config_path)
validate_config(config)
```

## Build Dry-Run Plan

```{r}
plan <- plan_conversion(
  config = config,
  input_dir = input_dir,
  participant_label = "01"
)

plan[, c("source_json", "status", "destination_dir", "destination_stem")]
```

## Convert

```{r}
conversion <- convert_plan(
  plan = plan,
  bids_dir = bids_dir,
  clobber = FALSE
)

conversion
```

## Summarize QA Report

```{r}
report <- summarize_plan(plan, conversion = conversion)
report$qa
```

## Inspect Outputs

```{r}
list.files(bids_dir, recursive = TRUE)
```

## CLI Equivalent

```{r, eval = FALSE}
# Create starter config
dcmtobids init-config --output ./config.json

# Preflight checks
dcmtobids doctor --skip-dcm2niix --config ./config.json --bids-dir ./bids

# Run conversion from helper-style files
dcmtobids run \
  --skip-dcm2niix \
  --dicom-dir ./input \
  --config ./config.json \
  --bids-dir ./bids \
  --participant 01 \
  --report ./report.json
```
